cardinalR: Generating interesting high-dimensional data structures

### 2025-09-05 paper-cardinalR.Rmd
ML changes up to linear branching section

There are numerous packages available in R for generating synthetic data, each designed with unique characteristics and focus areas. For example, `geozoo` (@barret2016) offers a large collection of geometric objects, allowing users to create and analyze specific shapes, primarily in lower-dimensional spaces. Another useful package is `snedata` (@james2025), which provides tools for generating simplified datasets useful for evaluating dimensionality reduction techniques like tSNE, often focusing on understanding and evaluating low-dimensional embeddings of complex data structures. Additionally, `mlbench` (@friedrich2024) includes a collection of well-known benchmark datasets commonly associated with established classification or regression challenges. In the field of single-cell omics, `splatter` (@luke2017) is particularly simulate complex biological data, effectively capturing nuances such as batch effects and differential expression.

There is a valuable opportunity to improve the generation of high-dimensional data structures by integrating geometric principles with advanced noise control and customizable clustering. The `geozoo` package provides a strong foundation but could be enhanced to support high-dimensional extensions with controlled noise and user-defined parameters for clustering. Similarly, while `snedata` focuses on abstract datasets for dimensionality reduction, adding features for generating high-dimensional data from geometric layouts would enhance its usability. The `mlbench` package could also benefit from allowing users to create datasets with specific geometric structures and noise profiles. Additionally, although `splatter` specializes in biological data simulation, it could be expanded to offer a broader framework for generating diverse geometric structures across dimensions, enabling detailed control over noise and clustering. Addressing these areas could lead to more robust high-dimensional data generation tools.

To address this gap, this paper introduces the `cardinalR` R package. This package provides a collection of functions designed to generate customizable data structures in any number of dimensions, starting from basic geometric shapes. `cardinalR` offers important functionalities that extend beyond the capabilities of existing tools, allowing users to: (i) construct high-dimensional datasets based on geometric shapes, including the option to enhance dimensionality by adding controlled noise dimensions; (ii) introduce adjustable levels of background noise to these structures; and (iii) create complex clustered data arrangements by using fundamental geometric forms, while maintaining their positions, scales, orientations, and sample sizes in arbitrary dimensional spaces. By providing these integrated features, `cardinalR` aims to provide researchers to generate more explainable and challenging synthetic datasets focused to the specific needs of evaluating algorithms in high-dimensions. This bridges the gap between geometric foundations and the flexible generation of complex synthetic data.

@@

A branching structure (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:branch)"} else { "Figure \\@ref(fig:branch-proj)"})`) captures trajectories that diverge or bifurcate from a common origin, similar processes such as cell differentiation in biology. We introduce a set of data generation functions specifically designed to simulate high-dimensional branching structures with various geometry, number of points, and the number of branches. Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:branching-tb-html)', '\\@ref(tab:branching-tb-pdf)'))` outlines these functions. The main arguments of the functions described in Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:arg-branching-tb-html)', '\\@ref(tab:arg-branching-tb-pdf)'))`.

@@

Each branch $i$ is constructed using $X_1 \sim U(-2, 2)$ and $X_2 = \exp(\pm s_iX_1) + \epsilon$, where: $\epsilon \sim U(0, 0.1)$ adds local jitter, $s_i \in [0.5, 2]$ is randomly sampled for each branch. The sign alternates between branches (odd: negative, even: positive exponent). This creates mirror-symmetric branches with different steepness and curvature where odd-numbered branches: decay pattern (reflected) and even-numbered branches: growth pattern. For $p > 2$, Gaussian noise $X_j \sim N(0, 0.1^2)$ is added to embed the $2\text{-}D$ branches into $p\text{-}D$, where $j = 3, \dots, p$.

@@

Each branch is a segment of a line with added jitter to simulate measurement noise. The branches differ in direction and location. Branches $1$ and $2$ are initialized with fixed slopes and intercepts. The Branch $1$ is generated from $X_1 \sim U(-2, 8)$, $X_2 = 0.5X_1 + \epsilon$, where $\epsilon \sim U(0, 0.5)$. The Branch $2$ is generated from $X_1 \sim U(-6, 2)$, $X_2 = -0.5X_1 + \epsilon$, where $\epsilon \sim U(0, 0.5)$. Branches $3$ to $k$ are added iteratively. Each additional branch $i$ starts at a location outside predefined exclusion zones to avoid overlap with the initial two branches. The $X_1$ values are defined over a short range, from $x_{start}$ to $x_{start} + 1$. The $X_2$ value is calculated using the formula $X_2 = s_i(X_1 - x_{start}) + y_{start} + \epsilon$, where $s_i$ is a chosen slope from a selected branch, and $\epsilon \sim U(0, 0.2)$. For $p > 2$, Gaussian noise $X_j \sim N(0, 0.05^2)$ is added to embed the $2\text{-}D$ branches into $p\text{-}D$, where $j = 3, \dots, p$.

>>>>

There are numerous packages available in R for generating synthetic data, each designed with unique characteristics and focus areas.The `geozoo` package (@barret2016) offers a large collection of geometric objects, allowing users to create and analyze specific shapes, primarily in lower-dimensional spaces. Another useful package is `snedata` (@james2025), which provides tools for generating simplified datasets useful for evaluating dimensionality reduction techniques like tSNE, often focusing on understanding and evaluating low-dimensional embeddings of complex data structures. Additionally, `mlbench` (@friedrich2024) includes a collection of well-known benchmark datasets commonly associated with established classification or regression challenges. In the field of single-cell omics, `splatter` (@luke2017) is designed to simulate complex biological data, capturing field-specific nuances such as batch effects and differential expression.

There is an  opportunity to improve the generation of high-dimensional data structures by integrating geometric principles with advanced noise control and customizable clustering. The `geozoo` package provides a strong foundation but could be enhanced to support high-dimensional extensions with controlled noise and user-defined parameters for clustering. Similarly, while `snedata` focuses on abstract datasets for dimensionality reduction, adding features for generating high-dimensional data from geometric layouts would enhance its usability. The `mlbench` package could also benefit from allowing users to create datasets with specific geometric structures and noise profiles. Additionally, although `splatter` specializes in biological data simulation, it could be expanded to offer a broader framework for generating diverse geometric structures across dimensions, enabling detailed control over noise and clustering. Addressing these areas could lead to more robust high-dimensional data generation tools.

To address this gap, this paper introduces the `cardinalR` R package. This package provides a collection of functions designed to generate customizable data structures in any number of dimensions, starting from basic geometric shapes. `cardinalR` offers important functionalities that extend beyond the capabilities of existing tools, allowing users to: (i) construct high-dimensional datasets based on geometric shapes, including the option to enhance dimensionality by adding controlled noise dimensions; (ii) introduce adjustable levels of background noise to these structures; and (iii) combine high-dimensional datasets into a single multi-faceted, clustered dataset in a space of arbitrary dimension. By providing these features, `cardinalR` provides researchers with a method to generate more explainable and challenging synthetic datasets focused to the specific needs of evaluating algorithms in high-dimensions. This bridges the gap between geometric foundations and the flexible generation of complex synthetic data.

@@

A branching structure (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:branch)"} else { "Figure \\@ref(fig:branch-proj)"})`) captures trajectories that diverge or bifurcate from a common origin, similar to processes such as cell differentiation in biology. We introduce a set of data generation functions specifically designed to simulate high-dimensional branching structures with various geometry, number of points, and the number of branches. Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:branching-tb-html)', '\\@ref(tab:branching-tb-pdf)'))` outlines these functions. The main arguments of the functions described in Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:arg-branching-tb-html)', '\\@ref(tab:arg-branching-tb-pdf)'))`.

@@

Each branch $i$ is constructed using $X_1 \sim U(-2, 2)$ and $X_2 = \exp(\pm s_iX_1) + \epsilon$, where: $\epsilon \sim U(0, 0.1)$ adds local jitter, $s_i \in [0.5, 2]$ is randomly sampled for each branch. The sign alternates between branches (odd: negative, even: positive exponent). This creates mirror-symmetric branches with different steepness and curvature where odd-numbered branches represent a decay pattern (reflected) and even-numbered branches represent a growth pattern. For $p > 2$, Gaussian noise $X_j \sim N(0, 0.1^2)$ is added to embed the $2\text{-}D$ branches into $p\text{-}D$, where $j = 3, \dots, p$.

@@

Each branch is a segment of a line with added jitter to simulate measurement noise. The branches differ in direction and location. Branches $1$ and $2$ are initialized with fixed slopes and intercepts. Branch $1$ is generated from $X_1 \sim U(-2, 8)$, $X_2 = 0.5X_1 + \epsilon$, where $\epsilon \sim U(0, 0.5)$. The Branch $2$ is generated from $X_1 \sim U(-6, 2)$, $X_2 = -0.5X_1 + \epsilon$, where $\epsilon \sim U(0, 0.5)$. Branches $3$ to $k$ are added iteratively. Each additional branch $i$ starts at a location outside predefined exclusion zones to avoid overlap with the initial two branches. The $X_1$ values are defined over a short range, from $x_{start}$ to $x_{start} + 1$. The $X_2$ value is calculated using the formula $X_2 = s_i(X_1 - x_{start}) + y_{start} + \epsilon$, where $s_i$ is a chosen slope from a selected branch, and $\epsilon \sim U(0, 0.2)$. For $p > 2$, Gaussian noise $X_j \sim N(0, 0.05^2)$ is added to embed the $2\text{-}D$ branches into $p\text{-}D$, where $j = 3, \dots, p$.
